---
title: "Final"
author: "Will Grimsich"
format: 
  pdf:
    mainfont: "Helvetica Neue"
    sansfont: "Arial"
    header-includes: |
      \usepackage{titling}
      \setlength{\droptitle}{-0.5in}
date: "June 9, 2026"
toc: TRUE
toc_depth: 2
execute:
  warning: FALSE
  message: FALSE
---

# [Github Repository](https://github.com/wgrimsich/ENVS-193DS_spring-2026_final)

# Set Up

```{r Loading Packages and Data}

library(tidyverse)
library(here)
library(janitor)
library(DHARMa)

nests <- read.csv(here("data","nest_data_final.csv"))
```

# Problems

## Problem 1. Research Writing

### a. Transparent Statistical Methods

In part 1, they used a **binomial "log odds" regression** to relate and predict (*Recurvirostra americana*) habitat occupation to distance from the water's edge.

In part 2, they used a **Chi-Squared test** to compare expected vs observed counts for habitat type and occupation by species.

### b. Suggestions for Rewriting

Utilizing binomial logistic regression analysis, we found a significant relationship between distance from the water's edge and American Avocet (*Recurvirostra americana*) microhabitat use.

Utilizing a Chi-Squared test, we found a significant relationship between bird species (American Avocet, Forster’s Tern, Black-necked Stilt) and habitat (managed pond, non-tidal marsh, salt pond, tidal marsh, or tidal mudflat).

### c. Further Analysis

One additional variable that could influence the probability of American Avocet (*Recurvirostra americana*) microhabitat use is **salinity**. This variable is continuous, numerical (ppm) and often calculated after measuring electrical conductivity in micro Siemens per centimeter (µS/cm). Salinity influences a plethora of biological interactions, specifically dehydration, [bodyweight and feeding/preening behavior.](https://bioone.org/journals/waterbirds/volume-26/issue-1/1524-4695_2003_026_0119_IOSOGA_2.0.CO_2/Impacts-of-Salinity-on-Growth-and-Behavior-of-American-Avocet/10.1675/1524-4695(2003)026%5B0119:IOSOGA%5D2.0.CO;2.full).

## Problem 2. Data Analysis

### a. Clean & Wrangle

```{r Cleaning Data}

nests_clean <- nests |> 
  select(case_control,
         ant_species,
         height_cm) |> 
  filter(ant_species %in% c("AB",
                            "RRB",
                            "BBR")) |> 
  mutate(ant_name = case_match(
    ant_species,
    "AB" ~ "C.sjostedti",
    "RRB" ~ "C.mimosae",
    "BBR" ~ "C.nigriceps"
  )) |> 
  mutate(ant_name = as_factor(ant_name),
         ant_name = fct_relevel(ant_name,
                                "C.sjostedti",
                                "C.mimosae",
                                "C.nigriceps"))

str(nests_clean)

slice_sample(nests_clean, n = 10)

```

### b. Response Variable

The 1s and 0s present within the "case_control" column of the data represent if a bird nest was observed (1) or not (0) within a specific tree.

### c. Purpose of Study

**Tree height** is a continuous, numerical variable measured in centimeters. As tree height increases, the probability of nest occurrence would increase because the tree contains more space for nesting and is further away from the ground and therefore potential predators. (Section 3: Results, ¶2)

**Acacia ant species** is a categorical, nominal variable. The four studied ant species differ in aggression of protecting their host trees, as aggressive protection increases, bird nesting also increases because of this inherent nest protection. (Section 3: Results, ¶3)

### d. Table of Models

| Model number | Predictor List           | Interaction Term? |
|--------------|:------------------------:|:-----------------:|
| 1            | Ant Species, Tree Height | No Interaction    |
| 2            | Ant Species, Tree Height | With Interaction  |

### e. Running the Models

```{r Creating Models, output=FALSE}

model1 <- glm(
  case_control ~ ant_name + height_cm,
  data = nests_clean,
  family = "binomial")

model1

model2 <- glm(
  case_control ~ ant_name*height_cm,
  data = nests_clean,
  family = "binomial")

model2

```

### f. Selecting the Best Model

```{r AIC}

AIC(model1,
    model2)

```

The best model as determined by Akaike's Information Criterion (AIC) was the model predicting the probability of bird nest occurrence as predicted by ant species interacting with tree height.

### g. Checking Diagnostics

```{r Plotting Diagnostics}

plot(
  simulateResiduals(model2)
)

```

### h. Visualizing Model Predictions

```{r GLM Plot}

```


### i. Writing a Caption for Figure



### j. Calculating Model Predictions



### k. Interpreting Results



## Problem 3. Affective & Exploratory Visualizations

### a. Comparing Visualizations



### b. Designing an Analysis



### c. Checking Assumptions & Running Analysis



### d. Creating Visualization



### e. Writing Caption



### f. Writing About Results



### g. Sharing Affective Visualization

